Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-05-26 manuscript/cpcrw_manuscript.Rmd
Incorporated VLB comments on May 19 draft

We sampled soils from a southeast slope (65.1620 °N, 147.4874 °W) at CPCRW, in a 60 m transition zone between lowland *Picea mariana* Mill BSP. and upland *Betula neoalaskana*, with significant white spruce (*Picea glauca*) presence. Stand density in this transition zone was 4060 ± 2310 trees ha^-1^, with basal area of 27.9 ± 7.0 m^2^ ha^-1^. The forest was at least 90 years old {cf. \\Morishita, 2014 #4316} according to tree rings taken at the stem base of several of the largest white spruce. The soil is characterized as a poorly-drained silt loam, and on average had ~20 cm of organic material over the mineral soil.

@@

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} quantified gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture, consistent with our results They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from *Picea mariana* sites near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C; these values are higher than we observed, consistent with the lower temperature range {Hamdi, 2013 #3664} and fundamental biokinetics {Davidson, 2006 #1669}. Waldrop et al. {, 2010 #4380} also observed flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites (but roughly similar to our observed CH~4~ emissions), and ~1-5 µg C-CO2 hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores. In an incubation of active-layer Alaskan permafrost peats, Treat et al. {, 2014 #4213} found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture. Finally, during the first 100 days of a year-long incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here.

@@

In contrast, the temperature sensitivities observed in this experiment were low (all less than 2.0, even when controlling for changes in soil moisture), but not unprecedented in comparison to a wide range of other laboratory soil incubations {Hamdi, 2013 #3664}. In contrast, observed surface CO~2~ fluxes at this range exhibited a Q~10~ of 5.1 ± 1.4 over a temperature range of 3.5-15 °C (personal communication, C. Anderson). It is important to note, however, that while increased temperature does not always drive C mineralization rates in forest mineral soils {Giardina, 2000 #2159}, it is linked with increases in soil moisture content can lead to changes in microbial community structure and GHG fluxes {Xue, 2016 #4361}. More surprisingly, Q~10~ values were lower in the drought treatment cores, a mathematical consequence of the fact that drought restricted CO~2~ respiration more at 4 than at 20 °C. **TODO: THOUGHTS ON THIS FROM A MICROBIAL PERSPECTIVE?**

@@

Somewhat unexpectedly, percent soil N was very significantly and positively correlated with both CO~2~ and CH~4~ fluxes (**Tables 1 and 2**). N interacts with microbial respiration via a number of complex, interactive, and still unclear mechanisms {Luo, 2006 #2443}, including reductions in belowground plant allocation, shifts in energy source or population of the saprotrophic community {Saiya-Cork, 2002 #3595} that leave it less capable of decomposing recalcitrant compounds , and perhaps abiotic stabliization mechanisms {Janssens, 2010 #2518}. Meta-analyses have generally shown negative to neutral effects of N deposition on microbial biomass {Treseder, 2008 #4421} and respiration {Ramirez, 2012 #4422}, and total soil respiration across ecosystems and biomes {Janssens, 2010 #2518}{Zhou, 2014 #3845}. These effect are likely due to several one or more mechanisms involving soil pH, ligninase enzymes, and phenol oxidase activity {Luo, 2006 #2443}.

@@

We focused on an experimental drought, rather than flooding, because of the well-drained nature of the field site: it is unlikely that the mid-slope forest we sampled in will ever suffer from thermokarst or excessive soil moisture, but too-dry conditions are a serious possibility in this low-precipitation ecosystem {Barber, 2000 #1320}. In addition, the soils here are not surface layer soils (where the majority of microbial activity and C mineralization of labile C takes place); taking them out of depth (where they are less exposed to O~2~, for example) may significantly change the abiotic conditions to which the microbial community is adapted. **TODO:MORE?**

@@

We are grateful to Jamie Hollingsworth for information about, and faciliating access to, the Caribou Poker Creeks Research Watershed Long-Term Ecological Research site. This research was supported by the Office of Science of the U.S. Department of Energy as part of the Terrestrial Ecosystem Sciences Program. The Pacific Northwest National Laboratory is operated for DOE by Battelle Memorial Institute under contract DE-AC05-76RL01830.

>>>>

We sampled soils from a southeast slope (65.1620 °N, 147.4874 °W) at CPCRW, in a 60 m transition zone between lowland *Picea mariana* and upland *Betula neoalaskana*, with significant white spruce (*Picea glauca*) presence as well. Stand density in this transition zone was 4060 ± 2310 trees ha^-1^, with basal area of 27.9 ± 7.0 m^2^ ha^-1^. The forest was at least 90 years old {cf. \\Morishita, 2014 #4316} according to tree rings taken at the stem base of several of the largest white spruce. The soil is characterized as a poorly-drained silt loam, and on average had ~20 cm of organic material over the mineral soil.

@@

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} quantified gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture, consistent with our results They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from *Picea mariana* sites near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C; these values are higher than we observed, consistent with the lower temperature range {Hamdi, 2013 #3664} and fundamental biokinetics {Davidson, 2006 #1669}. Waldrop et al. {, 2010 #4380} also observed flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites (but roughly similar to our observed CH~4~ emissions), and ~1-5 µg C-CO~2~ hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores. In an incubation of active-layer Alaskan permafrost peats, Treat et al. {, 2014 #4213} found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture. Finally, during the first 100 days of a year-long incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here.

@@

In contrast, the temperature sensitivities observed in this experiment were low (all less than 2.0, even when controlling for changes in soil moisture), but not unprecedented in comparison to a wide range of other laboratory soil incubations {Hamdi, 2013 #3664}. In contrast, observed surface CO~2~ fluxes at this range exhibited a Q~10~ of 5.1 ± 1.4 over a temperature range of 3.5-15 °C (personal communication, C. Anderson). It is important to note, however, that while increased temperature does not always drive C mineralization rates in forest mineral soils {Giardina, 2000 #2159}, it is linked with increases in soil moisture content can lead to changes in microbial community structure and GHG fluxes {Xue, 2016 #4361}. More surprisingly, Q~10~ values were lower in the drought treatment cores, a mathematical consequence of the fact that drought restricted CO~2~ respiration more at 4 than at 20 °C. There is evidence that climate warming changes the microbial decay dynamics of soil organic C compounds generally considered to be stable {Frey, 2013 #4062}{Bond-Lamberty, 2016 #4371}. Unlike surface soils, active layer soils, which store large quantities of soil C {Mueller, 2015 #4435}, are not subject to abundant inputs of fresh C from vegetation, so the starting quality of the native soil C in active layer soils may be older, more microbially processed, and dominated by more stable "heavy" organic C {Karlsson, 2011 #4436}. Thus, it is not surprising that these more stable C compounds would be metabolized by processes that have been reported to be less temperature-sensitive.

@@

Somewhat unexpectedly, percent soil N was very significantly and positively correlated with both CO~2~ and CH~4~ fluxes (**Tables 1 and 2**). N interacts with microbial respiration via a number of complex, interactive, and still unclear mechanisms {Luo, 2006 #2443}, including reductions in belowground plant allocation, shifts in energy source or population of the saprotrophic community {Saiya-Cork, 2002 #3595} that leave it less capable of decomposing recalcitrant compounds, and perhaps abiotic stabilization mechanisms {Janssens, 2010 #2518}. Meta-analyses have generally shown negative to neutral effects of N deposition on microbial biomass {Treseder, 2008 #4421} and respiration {Ramirez, 2012 #4422}, and total soil respiration across ecosystems and biomes {Janssens, 2010 #2518}{Zhou, 2014 #3845}. These effect are likely due to several one or more mechanisms involving soil pH, ligninase enzymes, and phenol oxidase activity {Luo, 2006 #2443}.

@@

We focused on an experimental drought, rather than flooding, because of the well-drained nature of the field site: it is unlikely that the mid-slope forest we sampled in will ever suffer from thermokarst or excessive soil moisture, but too-dry conditions are a serious possibility in this low-precipitation ecosystem {Barber, 2000 #1320}. In addition, the soils here are not surface layer soils (where the majority of microbial activity and C mineralization of labile C takes place); taking them out of depth (where they are less exposed to O~2~, for example) may significantly change the abiotic conditions to which the microbial community is adapted. However, focusing on the active layer provides crucial information about the potential loss of C from these soils, a risk that needs to be well understood as permafrost degradation leads to expansions in the depth of the active layer across the Arctic.

@@

We are grateful to Jamie Hollingsworth for information about, and facilitating access to, the Caribou Poker Creeks Research Watershed Long-Term Ecological Research site. This research was supported by the Office of Science of the U.S. Department of Energy as part of the Terrestrial Ecosystem Sciences Program. The Pacific Northwest National Laboratory is operated for DOE by Battelle Memorial Institute under contract DE-AC05-76RL01830.
